---
title: "Create initial distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{init_dist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MoDVE)
```

# Initial individuals and distribution

Among its inputs, `run_modve_sim()` requires a `data.frame` containing the 
attributes and distribution in the 3D matrix of the epiphytes that form the 
initial community.

Initial individuals can be sampled with `draw_initial_individuals()`, based on
species traits and the microhabitat. 
These vignette covers the usage and behaviour of `draw_initial_individuals()`.

# Input data

The sampling of initial individuals depends on species-level traits,
generated with `draw_species_traits()`, and the microhabitat matrix containing
the surface area of substrate available for epiphytes and light conditions, 
generated with `create_microhab_mat()`.

The assembly of these inputs is covered in their respective vignettes, so 
here we simply load them from saved data:

```{r load_inputs, eval=FALSE}
SpeciesPool <- read.csv("path_to_inputs/species_pool.csv")
Microhabitat <- readRDS("path_to_inputs/microhabitat_matrix.rds")
```

# Distribution parameters

`draw_initial_individuals` requires a list of parameters controlling how many 
individuals are sampled to initialise the community, containing at least four
elements:

```{r distr_param, eval=FALSE}
distr_params <- list(
  "IndividualsPerSpecies" = 23, # nb of individuals of each species
  "PercentageMaturePerSpecies" = 50, # how many % are initialised with mass below/above MassAtMaturity
  "ScalingPerHa" = FALSE, # is IndividualsPerSpecies a density per hectar, or a global number?
  "SurfaceBiomassScaling" = 100# global parameter g_S of the simulation,
    # controls how the surface area occupied by an epiphyte scales with its mass
)
```

In addition to `distr_params`, the function provides two more options 
controlling how individuals are allocated 

# Initial individual sampling

First, a number of individual equal to `IndividualsPerSpecies` 
(or `IndividualsPerSpecies * dimX * dimT / 10000` if `ScalingPerHa = TRUE`) 
is generated for each species.

The mass of the initial individuals is sampled uniformly between species traits 
`MassAtMaturity` and `MaxMass`, for `PercentageMaturePerSpecies`% of them, 
and between 0 and `MassAtMaturity` for the rest of them.

The age of individuals is set according to their mass and the species growth 
rate, and the surface area of substrate they require is obtained from
\deqn{S = M^{2/3} / g_S}.

Once the pool of initial individuals has been assembled, epiphytes of each 
species are distributed in the matrix among suitable voxels (that is, with light
conditions falling and sufficient surface area) and enough substrate surface
area.

Should the total surface area in the matrix be insufficient to fit all the 
initial individuals, the remaining individuals are added to the table as dead
individuals as the start of the simulation, with `Status = 2` and assigned to
voxel `{1, 1, 1}`.

`draw_initial_individuals()` offers two optional features for the allocation of 
individuals to voxels. First, if `largest_inds_first = TRUE`, individuals within
a species are allocated by order of mass, to represent a competitive advantage.
Second, `largest_voxels_first = TRUE` allocates individuals to voxels with the
most surface area first.

Note that species are processed sequentially, so the first species effectively
get priority.

# Usage

```{r call_draw_init_dist, eval=FALSE}
init_inds_tbl <- draw_initial_individuals(
  distr_params = distr_params, 
  species_df = SpeciesPool, 
  microhab_mat = Microhabitat, 
  path_to_output = NULL, # return the output table
  largest_inds_first = TRUE,
  most_surf_area_first = FALSE
)
```

The conformity of the resulting initial individuals and their distribution can 
be asserted with `check_init_dist()`, which is also useful if one considers 
importing the initial table from external sources:

```{r check_init_dist, eval=FALSE}
check_init_dist(
  InitDist = init_inds_tbl, 
  SpeciesPool = SpeciesPool,
  dimensions = dim(Microhabitat)
  )
```

# Example script

```{r create_init_dist, eval=FALSE}

distr_params <- parse_config("tests/config_a3.toml")
set.seed(distr_params$seed, kind="Mersenne-Twister")

# Prepare folder paths
DirectoryModelMain <- distr_params$DirectoryModelMain
dir_microhab <- distr_params$DirectoryMicrohabitatMain
DirectorySpeciesPoolsMain <- distr_params$DirectorySpeciesPools

dir.create(DirectoryModelMain, recursive=TRUE)

# Load microhabitat matrix
microhabitat_filename <- paste("MicrohabitatMatrix", distr_params$TimeStep, ".rds", sep="")
path_to_microhab <- file.path(dir_microhab, microhabitat_filename)
microhab_mat <- readRDS(path_to_microhab)
# Transform light values from relative to absolute
microhab_mat[, , , 3] <- microhab_mat[, , , 3] * distr_params$Imax

for (numPool in distr_params$numSpeciesPools[1]:distr_params$numSpeciesPools[2]) {

  #numPool <- 4
  #numReplicates <- 1

  # Load the species pool
  species_filename <- paste("SpeciesPool", numPool, ".csv", sep = "")
  Input_file <- file.path(DirectorySpeciesPoolsMain, species_filename)
  species_df <- read.csv(Input_file)

  for (numReplicates in seq_len(distr_params$replicatePerSpeciesPool)) {

    path_to_output <- file.path(DirectoryModelMain, paste(
      "ID_SpeciesP_", numPool, "_Rep_", numReplicates, ".csv", sep = ""
    ))

    largest_inds_first <- TRUE
    most_surf_area_first <- TRUE

    # Generate the distribution and individual traits
    draw_initial_individuals(
      distr_params, species_df, microhab_mat, path_to_output,
      largest_inds_first = distr_params$SingleSpeciesModel,
      most_surf_area_first = distr_params$MethodVoxel
    )
  }
}

```
