---
title: "Create species pool"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{species_pool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MoDVE)
```

# Species traits hyperparameters

One of the required inputs for running a `MoDVE` simulation is the `SpeciesPool`,
a table containing the growth, dispersal, reproduction and niche traits that 
define a species.

The values for these species traits can be imported from external sources,
or generated with `draw_species_traits()` from a list of species trait 
hyperparameters, `species_params`:

```{r hyperparams, eval=FALSE}
species_params <- list(
  # Growth traits
  MaxMassLogScaleRandom = FALSE,
  MaxMassRandom = c(2, 3000),
  MassAtMaturityRelativeRandom = c(0.5, 0.7),
  InterceptAgeMaturity = 2,
  ScalingAgeMaturity = 0.25,
  AgeAtMaturityDevCorr = 0.25,
  # Dispersal traits
  DispersalKernelRandom = c(0.03, 0.5),
  DispersalKernelAsymmetryRandom = c(0.5, 0.95),
  # Reproduction traits
  CorrelationMassRecruitment = TRUE,
  RecruitmentInvestmentRelMeanCorr = 0.1,
  RecruitmentInvestmentRelDevCorr = 0.25,
  RecruitmentInvestmentRelMeanRandom = c(0.07, 0,12),
  RecruitmentIncRandom = c(0, 0),
  # Light niche traits
  HeightBreadthRandom = c(0.15, 0.7),
  Imax = 900,
  kL = 0.6,
  LAI = 6
)
```

In this vignette, we walk through the different species traits, and how they
are sampled using the different elements of `species_params`. 

# Growth traits

Epiphytes can grow up to a maximum species-defined `MaximumMass`, and can 
disperse seeds after reaching a threshold mass `MassAtMaturity`.
Each year, assuming optimal light conditions (light == `OptimumLight`), 
individuals grow by an amount \eqn{\Delta M = K \times (M_{max} - M)}, 
where \eqn{K} is the growth rate.

The `MaximumMass` is generated from a uniform distribution with minimum and 
maximum`MaxMassRandom` (option `MaxMassLogScaleRandom = FALSE`), with an option to
transform the parameters with a log10 before sampling in order to skew the
distribution towards lower values (`MaxMassLogScaleRandom = TRUE`).

`MassAtMaturity` is set as a random fraction of `MaximumMass`, sampled in a
uniform with parameters `MassAtMaturityRelativeRandom`.

The growth rate is derived from first sampling the age at which epiphytes reach
maturity:
\deqn{Age_{mat} = InterceptAgeMaturity * M_{max}^{ScalingAgeMaturity} * U(1 \pm AgeAtMaturityDevCorr)}

The growth rate \eqn{K} is then derived from \eqn{Age_{mat}} so that it 
satisfies the growth rate above:
\deqn{K = -\rac{log(1 - \frac{M_{mat}}{M_{max}})} {Age_{mat}}}

# Dispersal traits

In `MoDVE`, reproduction and dispersal are coupled - only seedlings disperse.
Dispersal distance follows a negative exponential distribution with parameter
`DispersalKernel`. The resulting probability distribution is multiplied by a
gravity bias, with value `DispersalKernelAsymmetry` for all voxels below the
parent epiphyte and `1 - DispersalKernelAsymmetry` for all voxels above it or on
the same height.

Both parameters are sampled in uniform distributions, with respective
hyperparameters `DispersalKernelRandom` and `DispersalKernelAsymmetryRandom`.

 
# Reproduction traits

Fecundity \eqn{n_{R_{Pot}}}, the average number of seedlings produced by an epiphyte in one 
year (per square-meter of substrate area) is determined by 
\deqn{n_{R_{Pot}} = f(M) * Rec_{InvRel} * (1 + Rec_{Inc} * \frac{M - M_{mat}}{M_{max} - M_{mat}})}
where \eqn{f(M)} is a linear function of the individual's mass (independent of 
species), \eqn{Rec_{InvRel}} (`RecruitmentInvestmentRel`) is a species trait scaling represents the fraction 
of the epiphyte's biomass invested in reproduction, and \eqn{Rec_{Inc}} (`RecruitmentInc`) is a
species trait representing an increase in the reproduction allocation as the epiphyte
grows closer to its maximum mass.

The values for `RecruitmentInvestmentRel` and `RecruitmentInc` can be sampled in
two ways, depending on the value of `CorrelationMassRecruitment`:
- if `TRUE`, `RecruitmentInvestmentRel` is sampled in a uniform distribution
with min and max values `RecruitmentInvestmentRelMeanCorr` * (1 \eqn{\pm} `RecruitmentInvestmentRelDevCorr`), and `RecruitmentInc` is set to 0.
- if `FALSE`, `RecruitmentInvestmentRel` is sampled in a uniform distribution
with min and max values `RecruitmentInvestmentRelMeanRandom`
and `RecruitmentInc` is sampled in a uniform distribution with min and max
values `RecruitmentIncRandom.`

# Light niche traits

The light niche of the species is determined by first sampling the height niche,
the range of heights (relative to canopy height) in which the epiphyte species
can dwell.
Its center is sampled between 0 and 1, and its range is sampled in a uniform
with minimum and maximum values `HeightBreadthRandom`.

The minimum and maximum values of the height niche are then translated into 
minimum and maximum bounds of the light niche (resp `MinLight` and `MaxLight`:):
\deqn{I_{max} * e^{-k_L \times LAI (1 - h)}}
where `Imax` (light intensity above the canopy), `LAI` 
(leaf area index of the species) and `k_L` (light extinction coefficient) are 
parameters, and \eqn{h} is either the minimum or maximum height.
The `OptimumLight` trait is set as the average of `MinLight` and `MaxLight`, and
the three light parameters are then transformed into parameters `LightResponseA`,
`LightResponseB` and `LightResponseC` of the parabolic light response.

# Usage

Once the list `species_params` has been assembled, one can call the trait 
sampling function:

```{r call_draw_sp_traits, eval=FALSE}
species_params <- draw_species_traits(species_params)
```

This returns a list with sampled values for the traits for one species.
This function should be called multiple times and the output formatted as a
`data.frame` to form the final `SpeciesPool`:

```{r, secret_draw, include=FALSE}
species_params <- draw_rnd_species_params()
```

```{r assemble_sp_pool}
nb_species <- 5

library(purrr)

# Example with purr utilities
SpeciesPool <- map(rep(list(species_params), nb_species), draw_species_traits) |> 
  map(as.data.frame) |> 
  list_rbind()
SpeciesPool <- cbind("SpeciesID" = 1:nb_species, SpeciesPool)
SpeciesPool
```

Note that all species do not need to share the same `species_params` list.

# Example script

```{r create_sp_pool, eval=FALSE}

species_params <- parse_config("tests/config_a2.toml")
check_species_params(species_params)

# Prepare output table
SaveDirectory <- file.path(species_params$MainOutputDirectory)
dir.create(SaveDirectory, recursive=TRUE)
trait_names <- c("SpeciesID", species_trait_names())
sp_traits_df <- data.frame(matrix(ncol = length(trait_names), nrow = 0))
colnames(sp_traits_df) <- trait_names

set.seed(species_params$seed, kind="Mersenne-Twister")  # integer for fixed seed or NULL for random

for (pool_nb in 1:species_params$numSpeciesPools) {
  for (sp in 1:species_params$NumberOfSpecies) {

    species_traits <- c("SpeciesID" = sp, draw_species_traits(species_params))
    sp_traits_df[sp, ] <- as.data.frame(species_traits)
  }

  # Save trait dataframe
  SpeciesPoolFileName <- paste("SpeciesPool", pool_nb, ".csv", sep = "")
  write.csv(sp_traits_df, file.path(SaveDirectory, SpeciesPoolFileName),
            row.names = FALSE)
}

```
