---
title: "Create species pool"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{species_pool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MoDVE)
```

One of the required inputs for running a `MoDVE` simulation is the `SpeciesPool`,
a table containing the growth, dispersal, reproduction and niche traits that 
define a species.

# Growth traits

Epiphytes can grow up to a maximum species-defined `MaximumMass`, and can 
disperse seeds after reaching a threshold mass `MassAtMaturity`.
Each generation, assuming optimal light conditions, individuals grow by an 
amount \eqn{\Delta M = K \times (M_{max} - M)}, where \eqn{K} is the growth 
rate.

#'  * `MaximumMass` positive numeric, the maximum mass an individual can reach.
#'  * `MassAtMaturity` numeric between 0 and and `MaximumMass`,
#'  fraction of `MaximumMass` above which at individual can reproduce.
#'  * `GrowthRate` numeric between 0 an 1, the fraction of remaining growth an
#'  individual gains in a single generation
#'  (i.e, \eqn{\Delta M = K \times (M_{max} - M)} under optimal light conditions.

#'  * `MaxMassLogScaleRandom` logical. If `FALSE`, `MaxMass` is sampled in a
#'  uniform distribution with parameters `MaxMassRandom`. If `TRUE`, the value
#'  is instead sampled in a uniform of the log10's of `MaxMassRandom`, and the
#'  sampled value is then transformed back. This produces the same range of
#'  values, but skews the distribution towards the minimum.
#'  * `MaxMassRandom` a length-2 numeric vector containing the minimum and
#'  maximum MaxMass
#'  * `MassAtMaturityRelativeRandom` a length-2 numeric vector containing the
#'  minimum and maximum bounds (between 0 and 1) in which to sample what
#'  fraction of `MaxMass` the mass at maturity corresponds to.
#'
#'  * `InterceptAgeMaturity` numeric, the intercept of the function generating
#'  the age at maturity as a function of mass at maturity (see details)
#'  * `ScalingAgeMaturity` numeric between 0 and 1, the exponent of the function
#' generating the age at maturity as a function of mass at maturity
#'  * `AgeAtMaturityDevCorr` numeric, the range of possible deviation
#'  coefficients in the function generating age at maturity from mass at
#'  maturity.
#'

# Dispersal traits
 
# Reproduction traits

# Light niche traits

Also returns height niche, but not required (in fact not used) in the simulation.

# Usage

Once the list `species_params` has been assembled, one can call the trait 
sampling function:

```{r call_draw_sp_traits, eval=FALSE}
species_params <- draw_species_traits(species_params)
```

This returns a list with sampled values for the traits for one species.
This function should be called multiple times and the output formatted as a
`data.frame` to form the final `SpeciesPool`:

```{r, secret_draw, include=FALSE}
species_params <- draw_rnd_species_params()
```

```{r assemble_sp_pool, eval=FALSE}
nb_species <- 5

library(purrr)

# Example with purr utilities
map(rep(list(species_params), nb_species), draw_species_traits) |> 
  map(as.data.frame) |> 
  list_rbind()
```

Note that all species do not need to share the same `species_params` list.

# Example script

```{r create_sp_pool, eval=FALSE}

species_params <- parse_config("tests/config_a2.toml")
check_species_params(species_params)

# Prepare output table
SaveDirectory <- file.path(species_params$MainOutputDirectory)
dir.create(SaveDirectory, recursive=TRUE)
trait_names <- c("SpeciesID", species_trait_names())
sp_traits_df <- data.frame(matrix(ncol = length(trait_names), nrow = 0))
colnames(sp_traits_df) <- trait_names

set.seed(species_params$seed, kind="Mersenne-Twister")  # integer for fixed seed or NULL for random

for (pool_nb in 1:species_params$numSpeciesPools) {
  for (sp in 1:species_params$NumberOfSpecies) {

    species_traits <- c("SpeciesID" = sp, draw_species_traits(species_params))
    sp_traits_df[sp, ] <- as.data.frame(species_traits)
  }

  # Save trait dataframe
  SpeciesPoolFileName <- paste("SpeciesPool", pool_nb, ".csv", sep = "")
  write.csv(sp_traits_df, file.path(SaveDirectory, SpeciesPoolFileName),
            row.names = FALSE)
}

```
