---
title: "Generate microhabitat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{microhabitat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Work in progress, currently only used to hold the R code without shipping the 
package with R scripts.

```{r setup}
library(MoDVE)
```


```{r, create_mhb, eval=FALSE}
create_microhabitat_mat(
    config = config,
    shoot_dt = ShootsBegin,
    trunk_dt = TrunksBegin,
    vox_dt = vox_dt,
    path_to_output = path_to_output,
    dead_branches_id = dead_branches_id,
    dead_trees_id = dead_trees_id
    )
```


# Generating the microhabitat withouth MoF3D

You can, you will just need to make sure that your data conforms to the expected
format.

We have included functions that assert data is in the correct format

```{r no_mof3d, eval=FALSE}
check_microhabitat(Microhabitat)
``` 

The main `MoDVE` functions (`run_modve_sim()`) also call this function before
executing.


# Example script

The script below
uses input parameters in config and simulation data from Mof3d in the accompanying folder

```{r generate_microhab, eval=FALSE}
config <- parse_config("../tests/testthat/configs/config_a1.toml")

# Iteration controls
rep_forest <- config$ReplicateForest
time_start <- config$time_start
time_end <- config$time_end

# Folder structure
mof3d_res_dir <- file.path(config$DirectoryGroIMP, "Results")
mof3d_model_dir <- file.path(config$DirectoryGroIMP, "Model")
forest_global_file <- "Forest_param_global.txt"
forest_pass_file <- paste("Forest_param_pass", rep_forest, ".txt", sep = "")

# Set up output folder
output_dir <- file.path(config$DirectorySaveMain)
dir.create(output_dir, recursive = TRUE)
# Copy Mof3D forest data to result directory
file.path(mof3d_model_dir, forest_global_file) |>
  file.copy(to = output_dir, overwrite = FALSE)
file.path(mof3d_model_dir, forest_pass_file) |>
  file.copy(to = output_dir, overwrite = FALSE)

# Load dimensions of forest patch from global forest file
path_to_forest_global <- file.path(mof3d_model_dir, forest_global_file)
GlobalForest <- read.table(path_to_forest_global, sep = "\t", row.names = 1)
config$MaxX <- GlobalForest["MaxX", 1]
config$MaxY <- GlobalForest["MaxY", 1]
config$MaxZ <- GlobalForest["MaxZ", 1]
config$corridor <- GlobalForest["WidthCorridor", 1]

shootFile <- paste("shoots_replicate_", rep_forest, "_time_step_", sep = "")
trunkFile <- paste("trees_replicate_", rep_forest, "_time_step_", sep = "")
voxFile <- paste("voxel_replicate_", rep_forest, "_time_step_", sep = "")
vox_colnames <- c("x", "y", "z", "leafarea")

# Load data for first time step
ShootsBegin <- file.path(mof3d_res_dir, paste(shootFile, time_start, ".txt", sep = "")) |>
  read.table(sep = "\t", header = TRUE, skip = 1)
TrunksBegin <-   file.path(mof3d_res_dir, paste(trunkFile, time_start, ".txt", sep = "")) |>
  read.table(sep = "\t", header = TRUE, skip = 8)

if (!config$hasDynamicMicrohabitat) time_end <- time_start

for (i in time_start:time_end) {

  print(paste("Time step", i))

  # Load shoot and trunk files of actual and next timestep: Shoots at begin of year
  # and at the end of year/begin of next year
  # Previous time step
  if (i != time_start) {
    # Last step end becomes beginning of this step
    ShootsBegin <- ShootsEnd
    TrunksBegin <- TrunksEnd
  }
  ShootsEnd <- file.path(mof3d_res_dir, paste(shootFile, i + 1, ".txt", sep = "")) |>
    read.table(sep = "\t", header = TRUE, skip = 1)
  TrunksEnd <- file.path(mof3d_res_dir, paste(trunkFile, i + 1, ".txt", sep = "")) |>
    read.table(sep = "\t", header = TRUE, skip = 8)

  # Find which branch segments and trunks die this time step
  dead_branches_id <- find_dead_segments(ShootsBegin, ShootsEnd)
  dead_trees_id <- find_dead_trees(TrunksBegin, TrunksEnd)

  # Load file containing information about leaf area per voxel
  if (config$calcLightConditions) {
    vox_dt <- read.table(
      file.path(mof3d_res_dir, paste(voxFile, i, ".txt", sep = "")), 
      sep = "\t", 
      header = FALSE, 
      skip = 2,
      col.names = append(vox_colnames, "empty_column") # issue with reading cols
      )[vox_colnames]
    
    # Convert indexing from Java (first = 0) to R indexing convention (first = 1)
    Voxels[, 1:3] <- Voxels[, 1:3] + 1
    
  } else vox_dt <- NULL

  path_to_output <- file.path(
    output_dir, paste("MicrohabitatMatrix_new", i, ".rds", sep = "")
  )

  create_microhabitat_mat(
    config = config,
    shoot_dt = ShootsBegin,
    trunk_dt = TrunksBegin,
    vox_dt = vox_dt,
    path_to_output = path_to_output,
    dead_branches_id = dead_branches_id,
    dead_trees_id = dead_trees_id
    )
}

```
