---
title: "Generate microhabitat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{microhabitat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MoDVE)
```

# The microhabitat matrix in MoDVE

To run the simulation, MoDVE requires a 3D matrix (the microhabitat) containing
three layers of data, for each voxel:
- the amount of surface area available for epiphytes to grow on. Surface area
comprises the surface of trunks and the upper half of all branch segments of
all trees in the plot. It determines the total mass (and therefore, number) of
epiphytes that the voxel can support. This influences fecundity and mortality,
through the carrying capacity.
- the amount of surface area that is going to be lost in this generation due
to branch fall and tree mortality. 
Surface area loss is one of the sources of epiphyte mortality.
- the intensity of light in the voxel, determined by shading from the leaf mass
above the voxel and horizontal reflectance from nearby leaf compartments.
This influences fecundity, growth, and epiphytes die when light conditions fall
outside of their light niche.

# Required inputs

The matrix will typically be assembled from the output of MoF3D, a tropical 
forest functional-structural tree model (FSTM), but see the next section for
alternative options.

MoF3D simulates realistic 3D tree shapes and leaf biomass of a tropical forest 
stand, given a set of functional and structural traits.

Among its output, MoF3D produces three text tables that are of interest for 
deriving the microhabitat matrix:
- `trunk_dt` (MoF3D option `ReportTrees`): contains the (x and y) coordinates
and geometric attributes (height and diameter) for the trunk of each tree.
- `shoot_dt` (MoF3D option `ReportShoots`): contains the coordinates 
(x, y, z for the start and end of the segment) and geometric attributes 
(length, diameter, branch order) for each branch segment.
- `vox_dt` (MoF3D option `ReportVoxel`): contains the total amount of leaf 
biomass in each voxel.

```{r load_mof3d_inputs, eval=FALSE}
trunk_dt <- read.table("trees_replicate_0_timestep_1.txt", sep = "\t", header = TRUE, skip = 8)
shoot_dt <- read.table("shoots_replicate_0_timestep_1.txt", sep = "\t", header = TRUE, skip = 1)
```

In order to derive the amount of surface area loss, if applicable, `MoDVE` also
requires declaring which trees and branch segments are lost between the present
time step and the next one. 
This can be supplied manually as a vector of IDs, but the package also provides
utilities to identify the dying trees and branch segments automatically:

```{r find_dead_segments, eval=FALSE}
trunk_dt_next <- read.table("trees_replicate_0_timestep_2.txt", sep = "\t", header = TRUE, skip = 8)
shoot_dt_next <- read.table("shoots_replicate_0_timestep_2.txt", sep = "\t", header = TRUE, skip = 1)
dead_branches_id <- find_dead_segments(shoot_dt, shoot_dt_next)
dead_trees_id <- find_dead_trees(trunk_dt, trunk_dt_next)
```

# Input parameters

In addition to the MoF3D output data, a list `config` of parameters is required.
The list must contain at least the following elements:

1. `MaxX`, `MaxY`, `MaxZ` specifying the dimensions of the forest landscape.
`corridor`, if used in the MoF3D simulation, specifies the size (in voxels) of 
a tree-free corridor around the edges of the forest area, in which branches can
grow into, but not tree trunks.

2. `calcSurfaceArea`, `calcSurfaceAreaLoss` and `calcLightConditions` are all
boolean conditions enabling the generation of each microhabitat layer. While
all three layers are required to run the simulation, it may be desirable to
re-calculate only a subset of the layers.

3. `kL` and `DistVoxToConsider` are parameters controlling light extinction and
diffusion, respectively. `kL` parametrises the Beer-Lambert light extinction 
function:
\deqn{e^{-/frac{k_L * \sum^z LA_z}{10^4}}
where \eqn{\sum^z LA_z} denotes the total leaf area in all voxels above a
given voxel.
`DistVoxToConsider` sets the range of neighbouring voxels which contribute to
the light condition in the voxel of interest through reflectance.

# Generating the microhabitat matrix from MoF3D output

Once the required inputs have been assembled, assembling the matrix is 
straightforward:

```{r, create_mhb, eval=FALSE}
create_microhabitat_mat(
    config = config,
    shoot_dt = shoot_dt,
    trunk_dt = trunk_dt,
    vox_dt = vox_dt,
    path_to_output = path_to_output,
    dead_branches_id = dead_branches_id, # NULL if no loss
    dead_trees_id = dead_trees_id # NULL if no loss
    )
```


# Generating the microhabitat withouth MoF3D

It is possible to import data from another FSTM to use as input for 
`create_microhabitat_mat()`, as long as the data satisfies
the required input structure, which can be verified with helper functions:

```{r check_tbl, eval=FALSE}
check_shoot_dt(shoot_dt)
check_trunk_dt(trunk_dt)
check_vox_dt(vox_dt)
```

Alternatively, the microhabitat matrix can be assembled through other means.
In order to be used as input for the main simulation (`run_modve_sim()`) and other 
input preparation functions, it will need to pass a format check:

```{r no_mof3d, eval=FALSE}
check_microhabitat(Microhabitat)
``` 

# Example script

The script below uses input parameters in config and simulation data from Mof3d in the accompanying folder

```{r generate_microhab, eval=FALSE}
config <- parse_config("../tests/testthat/configs/config_a1.toml")

# Iteration controls
rep_forest <- config$ReplicateForest
time_start <- config$time_start
time_end <- config$time_end

# Folder structure
mof3d_res_dir <- file.path(config$DirectoryGroIMP, "Results")
mof3d_model_dir <- file.path(config$DirectoryGroIMP, "Model")
forest_global_file <- "Forest_param_global.txt"
forest_pass_file <- paste("Forest_param_pass", rep_forest, ".txt", sep = "")

# Set up output folder
output_dir <- file.path(config$DirectorySaveMain)
dir.create(output_dir, recursive = TRUE)
# Copy Mof3D forest data to result directory
file.path(mof3d_model_dir, forest_global_file) |>
  file.copy(to = output_dir, overwrite = FALSE)
file.path(mof3d_model_dir, forest_pass_file) |>
  file.copy(to = output_dir, overwrite = FALSE)

# Load dimensions of forest patch from global forest file
path_to_forest_global <- file.path(mof3d_model_dir, forest_global_file)
GlobalForest <- read.table(path_to_forest_global, sep = "\t", row.names = 1)
config$MaxX <- GlobalForest["MaxX", 1]
config$MaxY <- GlobalForest["MaxY", 1]
config$MaxZ <- GlobalForest["MaxZ", 1]
config$corridor <- GlobalForest["WidthCorridor", 1]

shootFile <- paste("shoots_replicate_", rep_forest, "_time_step_", sep = "")
trunkFile <- paste("trees_replicate_", rep_forest, "_time_step_", sep = "")
voxFile <- paste("voxel_replicate_", rep_forest, "_time_step_", sep = "")
vox_colnames <- c("x", "y", "z", "leafarea")

# Load data for first time step
ShootsBegin <- file.path(mof3d_res_dir, paste(shootFile, time_start, ".txt", sep = "")) |>
  read.table(sep = "\t", header = TRUE, skip = 1)
TrunksBegin <-   file.path(mof3d_res_dir, paste(trunkFile, time_start, ".txt", sep = "")) |>
  read.table(sep = "\t", header = TRUE, skip = 8)

if (!config$hasDynamicMicrohabitat) time_end <- time_start

for (i in time_start:time_end) {

  print(paste("Time step", i))

  # Load shoot and trunk files of actual and next timestep: Shoots at begin of year
  # and at the end of year/begin of next year
  # Previous time step
  if (i != time_start) {
    # Last step end becomes beginning of this step
    ShootsBegin <- ShootsEnd
    TrunksBegin <- TrunksEnd
  }
  ShootsEnd <- file.path(mof3d_res_dir, paste(shootFile, i + 1, ".txt", sep = "")) |>
    read.table(sep = "\t", header = TRUE, skip = 1)
  TrunksEnd <- file.path(mof3d_res_dir, paste(trunkFile, i + 1, ".txt", sep = "")) |>
    read.table(sep = "\t", header = TRUE, skip = 8)

  # Find which branch segments and trunks die this time step
  dead_branches_id <- find_dead_segments(ShootsBegin, ShootsEnd)
  dead_trees_id <- find_dead_trees(TrunksBegin, TrunksEnd)

  # Load file containing information about leaf area per voxel
  if (config$calcLightConditions) {
    vox_dt <- read.table(
      file.path(mof3d_res_dir, paste(voxFile, i, ".txt", sep = "")), 
      sep = "\t", 
      header = FALSE, 
      skip = 2,
      col.names = append(vox_colnames, "empty_column") # issue with reading cols
      )[vox_colnames]
    
    # Convert indexing from Java (first = 0) to R indexing convention (first = 1)
    Voxels[, 1:3] <- Voxels[, 1:3] + 1
    
  } else vox_dt <- NULL

  path_to_output <- file.path(
    output_dir, paste("MicrohabitatMatrix_new", i, ".rds", sep = "")
  )

  create_microhabitat_mat(
    config = config,
    shoot_dt = ShootsBegin,
    trunk_dt = TrunksBegin,
    vox_dt = vox_dt,
    path_to_output = path_to_output,
    dead_branches_id = dead_branches_id,
    dead_trees_id = dead_trees_id
    )
}

```
