---
title: "Run the simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
#options(warn=-1)  # Suppress warnings
#options(digits.secs=3)  # 3 decimal digits for seconds
```

```{r setup, eval=FALSE}
library(MoDVE)
```

# Global simulation parameters

`run_modve_sim()` requires the following list of simulation controls:

```{r sim_params}
sim_params <- list(
  # Simulation loop controls
  "InitialTimeStep" = 1, # When does the simulation start ?
  "timeSteps" = 5, # For how many years should the simulation run?
  "StopCriterionHa" = 3 * 10000, # will stop if global density per hectare goes above this
  # Environmental controls
  "hasDynamicMicrohabitat" = FALSE, # is the forest static or dynamic?
  "Imax" = 500, # Light intensity above the canopy
  "SurfaceBiomassScaling" = 100, # required S.A. = mass^(2/3) / SurfaceBiomassScaling
  # Reproduction parameters
  "InterceptRecruitment" = 30,  # intercept of the mass-to-fecundity function
  "SlopeRecruitment" = 0.1, # slope of the mass-to-fecundity function
  # Competition parameters
  "massDepCompetition" = TRUE, # do largest individuals win competition?
  # Mortality parameters
  "use_mass_dep_mortality" = FALSE, 
  # Only required if use_mass_dep_mortality = FALSE
  "MortRateRandom" = 0.1, # base mortality rate
  # Only required if use_mass_dep_mortality = TRUE
  "MortRateMass" = NULL,
  "MortRateMassScaling" = NULL
)
```

More details on the simulation parameters can be found on the help page (`?run_modve_sim`) and in the user guide.

The validity of the parameters can be checked with `check_sim_params()`, which is called before the simulation.

# Inputs

In addition to the simulation parameters, the MoDVE simulation requires three inputs. The preparation of each of these inputs is the subject of a vignette. If you haven't done so already, please go through them first before tackling the simulation.

| Input | Description | Vignette | Check |
|----|----|----|----|
| `Microhabitat` | Generation of the microhabitat matrix from MoF3D output | `vignette("")` | `check_microhabitat()` |
| `SpeciesPool` | Generation of the species traits | `vignette("")` | `check_species_df()` |
| `InitDist` | Generation of the initial individuals and their distribution | `vignette("")` | `check_init_dist()` |

Alternatively, one can also provide inputs generated from other sources, provided they conform to the data format expected by the simulation (this can be verified with the `check_*()` functions).

For this vignette, we assume these inputs have already been generated and we simply load them in the environment:

```{r load_inputs,  eval=FALSE}
path_to_microhabitat <- "my_data/microhabitat.csv"
Microhabitat <- utils::read.csv(path_to_microhabitat, sep = ",", header = TRUE)

```

For convenience, `run_modve_sim()` also accepts paths to CSV tables and loads them before the simulation, which may sometimes be preferable to passing large data frames.

In the case of dynamic forest landscapes `sim_params$hasDynamicMicrohabitat = TRUE`), `Microhabitat` must be a vector of either paths or `data.frame`s with one element per time step.

# Running the simulation

Once the inputs and parameters have been assembled, running the model is straightforward:

```{r run_model, eval=FALSE}

  run_modve_sim(
    sim_params = sim_params,
    SpeciesPool = SpeciesPoolFileName,
    Microhabitat = Microhabitat,
    InitDist = path_to_init_file,
    path_to_ind_output = "my_data/outputs/individual_output.csv",
    path_to_sp_output = "my_data/outputs/species_output.csv",
    path_to_comm_output = "my_data/outputs/community_output.csv"
  )
```

# Outputs

The output is written once per year. For individual output, the simulation loop saves a separate file for each generation (by inserting the generation number at the end of `path_to_ind_output()`), while the species and community outputs are overwritten with the last year.

# Example simulation script

The script below runs the simulation across multiple sets of species traits, with multiple replicates for each trait set, and includes RNG management and parallelisation of the simulation.

```{r example, eval=FALSE}

# Parse input configuration file
config <- parse_config("tests/config_a4.toml")

# Output directory
dir_output <- config$DirectoryModelResults
dir.create(dir_output, recursive = TRUE)

# Detect the number of CPU cores.
# On a HPC node, use the $SLURM_NTASKS env var if defined.
nTasks <- Sys.getenv("SLURM_NTASKS")
if (nTasks != "") { 
  numCores <- strtoi(nTasks) 
  } else { 
  numCores <- parallel::detectCores() 
  }
doParallel::registerDoParallel(numCores)

# RNG seed
random_state_file <- config$RandomState
if (!is.null(random_state_file) && file.exists(random_state_file)) {
  writeLines("Loading previous random number generator state from ", random_state_file)
  restore_rng(random_state_file)
} else {
  set.seed(config$seed, kind = "Mersenne-Twister")
}
# Save current random state.
writeLines("Storing the random number generator state...")
save_rng(file.path(dir_output, "random_state_seed.RData"))

# Main loop for the community model for each species pool and for each replicate
# Replicates use different RNG streams, but the same seed, so reproducible even when parallelised.
pairs <- expand.grid(config$numSpeciesPools, config$replicatePerSpeciesPool)
colnames(pairs) <- c("numPool", "r")
output <- foreach::foreach(pair_idx = seq_len(nrow(pairs))) %dorng% {

  numPool <- pairs$numPool[pair_idx]
  r <- pairs$r[pair_idx]

  # Load initial distribution of individuals
  init_filename <- paste("ID_SpeciesP_", numPool, "_Rep_", r, ".csv", sep = "")
  path_to_init_file <- file.path(config$DirectoryModelMain, init_filename)

  # Load species pool
  sp_filename <- paste("SpeciesPool", numPool, ".csv", sep = "")
  SpeciesPoolFileName <- file.path(config$DirectorySpeciesPools, sp_filename)

  # Load microhabitat (static forest)
  time_seq <- config$InitialTimeStep + seq_len(config$timeSteps) - 1
  path_to_microhab <- paste("MicrohabitatMatrix", time_seq, ".rds", sep="")
  Microhabitat <- file.path(config$DirectoryMicrohabitat, path_to_microhab)

  # Create Save-Directory for each each replicate/initialDistribution
  dir_output_this_run <- file.path(dir_output, paste("ID_SpeciesP_", numPool, "_Rep_", r, sep=""))
  dir.create(dir_output_this_run, recursive = TRUE)
  path_to_ind_output <- file.path(dir_output_this_run, "IndividualMatrixTimeStep.csv")
  path_to_sp_output <- file.path(dir_output_this_run, "SpeciesSummary.csv")
  path_to_comm_output <- file.path(dir_output_this_run, "CommunitySummary.csv")

  # Run the IBM
  run_modve_sim(
    sim_params = config,
    SpeciesPool = SpeciesPoolFileName,
    Microhabitat = Microhabitat,
    InitDist = path_to_init_file,
    path_to_ind_output,
    path_to_sp_output,
    path_to_comm_output
  )
  return(NULL)
}


```
